Portable Multi- and Many-Core Performance for Finite-Difference/Element Codes; Application to the Free-Surface Component of NEMO (NEMOLite2D 1.0).

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt

The authors thank the reviewers for their detailed feedback on our paper.
We address the points raised below, beginning with the first (anonymous)
referee and then moving on to those from Carlos Osuna.

Anonymous referee:

1/ The reviewer is right to be concerned with seeing how much
   performance may be lost by the re-structuring required for
   PSyKAl. However, the original form of the NEMOLite2D code that we
   started with for this work was unoptimised. This is emphasised by
   the results in Figure 4 which show that, for the majority of cases,
   the final PSyKAl version was significantly faster. In contrast, our
   previous paper (Porter et al, 2016) tackled the "Shallow" code
   which has had optimisations applied to it over a period of some
   twenty years and was therefore a much more demanding test. There,
   the initial PSyKAl re-structuring did significantly harm
   performance. However, we showed that it was possible to recover the
   performance of the optimised original by applying code
   transformations that were PSyKAl-compliant. We therefore do not
   think that showing results for the optimisations
   applied to the original form of NEMOLite2D is necessary.

2/ The PSyKAl approach is in fact applicable to both Finite Element and
   Finite Difference codes (and in fact, LFRic, the largest project making use
   of it is using Finite Elements). We have updated the title to make
   this clear.

3/ A new section, "Related Approaches", has been added to the paper. In this
   we bring together a comparison of various other approaches including OCCA.
   This makes explicit the differences and makes clearer the advantages of
   PSyKAl.

4/ We also use the new "Related Approaches" section to emphasise that
    although the tool to generate the PSy layer exists ("PSyclone"),
    the current paper is about code structuring.

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt

2/ The PSyKAl approach is in fact applicable to both Finite Element and
   Finite Difference codes (and in fact, LFRic, the largest project making use
   of it is using Finite Elements). We have updated the title to make
   this clear.

3/ A new section, "Related Approaches", has been added to the paper. In this
   we bring together a comparison of various other approaches including OCCA.
   This makes explicit the differences and makes clearer the advantages of
   PSyKAl.

4/ We also use the new "Related Approaches" section to emphasise that
    although the tool to generate the PSy layer exists ("PSyclone"),
    the current paper is about code structuring.

Carlos Osuna's general review:

We have added a few key lines of source code from the Momentum kernel in
section 3.3. We have also added text that explains that the apparently poor
performance of this kernel is down to the number of division operations.

The suggestion of a "Related Work" section is a good one and we have added
this just after the Introduction.

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: add mom kernel fragment

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. Without this
information we have no way of knowing whether further optimisations
might yield worthwhile performance improvements. Therefore, we
consider the performance of the original, serial NEMOLIte2D code on
the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
which provides a relatively simple way of characterising the
performance of a code in terms of whether it is memory-bandwidth bound
or compute bound. To do so we follow the approach suggested
in~\cite{para_pearls} and construct a roofline model using the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
to obtain appropriate upper bounds on the memory bandwidth and
floating-point operations per second (FLOPS), respectively. Since we
are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is well below the peak performance of the
CPU, this is due to the balance of floating-point operations that it
contains. Improving this is outside the scope of the current work
since here we are focused on the {PS}y{KA}l separation of concerns and
the introduction of parallelism in the PSy layer.

>>>>

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. Without this
information we have no way of knowing whether further optimisations
might yield worthwhile performance improvements. Therefore, we
consider the performance of the original, serial NEMOLite2D code on
the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
which provides a relatively simple way of characterising the
performance of a code in terms of whether it is memory-bandwidth bound
or compute bound. To do so we follow the approach suggested
in~\cite{para_pearls} and construct a roofline model using the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
to obtain appropriate upper bounds on the memory bandwidth and
floating-point operations per second (FLOPS), respectively. Since we
are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is well below the peak performance of the
CPU, this is due to the balance of floating-point operations that it
contains and in particular, the number of division
operations. (Division costs at least eight times as much as a
multiplication in the Ivy Bridge core~\citep{fog_instructions}.) For
instance, a fragment of the most costly part of the momentum kernel is
shown below:
\begin{verbatim}
    ...
    ! -pressure gradient
    hpg = -g * (hu(ji,jj) + sshn_u(ji,jj)) * e2u(ji,jj) * &
           (sshn(ji+1,jj) - sshn(ji,jj))

    ! -linear bottom friction (implemented implicitly.)
    ua(ji,jj) = (un(ji,jj) * (hu(ji,jj) + sshn_u(ji,jj)) + rdt * &
                 (adv + vis + cor + hpg) / e12u(ji,jj)) / &
                (hu(ji,jj) + ssha_u(ji,jj)) / (1.0_wp + cbfr * rdt) 
\end{verbatim}
The liberal use of the division operation is clearly something that
can be improved upon. However, this is outside the scope of the
current work since here we are focused on the {PS}y{KA}l separation of
concerns and the introduction of parallelism in the PSy layer.
